Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer

### 2021-01-24 manuscript/secs/results.tex
few minor changes

Out of 150 HERV loci tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (\emph{P} adjusted < 0.05) (Figure \ref{fig:dge_res}a,b, red points).
The larger number of over-expressed genes in the comparison with healthy tissue could be consistent with field effect bias but could also be due to a larger healthy tissue sample size and concomitant increased statistical power.
We did not investigate this point further because the intersection of the tumor specific proviruses from both comparisons yielded 11 HERV loci clearly unaffected by field effect bias and expressed in a tumor specific pattern (Figure \ref{fig:dge_res}c).
These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of immunogenic peptides, so we sought to further characterize their expression patterns and validate their tumor specificity in an independent cohort.

First, to confirm the ability of the 11 HERVs to distinguish tumor from non-tumor tissue, we used their expression level to classify samples in the discovery cohort.
As expected, when tested on the same samples from which their tumor specificity was discovered, most of the HERVs had excellent performance as biomarkers (Supplementary Figure 2).
Second, to determine whether the age difference across phenotypes was likely to have artificially inflated the apparent tumor specificity of the markers, we measured the correlation of HERV expression level with subject age.
Although a weak positive correlation was apparent (Supplementary Figure 3a), comparison of test statistics from linear models predicting expression level from age and phenotype showed a negligible effect of age (Supplementary Figure 3b).
Third, to assess the relationship between HERV expression and tumor characteristics, we plotted HERV expression separately for each category of tumor microsatellite instability status and stage.
The tumor specific HERVs tended to be higher in microsatellite stable tumors with the notable exceptions of HERV-H Xp22.32 and HERV-H 20p11.23 (Supplementary Figure 4a).
No trend in expression level relative to tumor stage was apparent (Supplementary Figure 4b).
Finally, to determine the relationship between HERV expression and cytotoxic activity within tumors, we measured the correlation between HERV expression and the expression of six markers of cytotoxic activity.
The correlation was negligible for most HERVs, but HERV1-I 3p22.3 had a weakly negative association (Supplementary Figure 5).

To verify the consistency of associations between proviral expression and tissue phenotype and to validate the utility of tumor specific HERVs as potential biomarkers, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 276 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 208 tumor samples (Supplementary Table 1).
Of the 150 HERVs moderately expressed (i.e. expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in cohort \#1.
Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in cohort \#1 (Supplementary Figure 1c).
As tumor classifiers in this cohort, most of the 11 tumor specific HERVs had weaker performance (Supplementary Figure 6).
However, HERV-H Xp22.32 and HERV1-I 3p22.3 maintained excellent discriminatory ability.
Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated as consistently tumor specific in cohort \#1 (Table \ref{tab:val}).
Note that nominally significant (i.e. \emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered significant for the purpose of validation due to the small number of healthy samples.

To confirm the accuracy of our bioinformatic and statistical methods, additional validation testing was performed on a third set of samples.
Testing in the discovery cohort relied on latent factor analysis to adjust for confounding factors across tissue phenotypes.
An alternative approach is to compare pairs of samples matched by subject.
This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for demographic, technical, and other possible confounders more effectively.
Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples were highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1d).
The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.

\subsection*{Proviruses are expressed in healthy colorectal tissue}
Several HERV proviruses were expressed in tumor specific patterns, in accord with our hypothesis.
However, a surprising finding of our survey of unannotated retroviral transcripts in colorectal transcriptomes was the unexpectedly high level of expression of HERV elements in healthy samples.
The number of differentially expressed genes (\emph{P} adjusted < 0.05) with higher expression in healthy and tumor adjacent samples relative to tumor samples was 94 and 13, respectively.
For comparison, the number of differentially expressed genes with lower expression in healthy and tumor adjacent samples relative to tumor samples was 28 and 14, respectively.
This result is demonstrated in Figure \ref{fig:dge_res}a,b (blue points).
A high number of differentially expressed genes with over-expression in healthy tissue was also found in cohort \#1 (data not shown).

>>>>

Out of 150 HERV loci tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (\emph{P} adjusted < 0.05) (Figure \ref{fig:dge_res}a,b, red points).
The larger number of over-expressed genes in the comparison with healthy tissue could be consistent with field effect bias but could also be due to a larger healthy tissue sample size and concomitant increased statistical power.
We did not investigate this point further because the intersection of the tumor specific proviruses from both comparisons yielded 11 HERV loci clearly unaffected by field effect bias and expressed in a tumor specific pattern (Figure \ref{fig:dge_res}c).
These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of immunogenic peptides, so we sought to further characterize their expression patterns in the discovery cohort.

First, to confirm the ability of the 11 HERVs to distinguish tumor from non-tumor tissue, we used their expression level to classify samples.
As expected, when tested on the same samples from which their tumor specificity was discovered, most of the HERVs had excellent performance as biomarkers (Supplementary Figure 2).
Second, to determine whether the age difference across phenotypes was likely to have artificially inflated the apparent tumor specificity of the markers, we measured the correlation of HERV expression level with subject age.
Although a weak positive correlation was apparent (Supplementary Figure 3a), comparison of test statistics from linear models predicting expression level from age and phenotype showed a negligible effect of age (Supplementary Figure 3b).
Third, to assess the relationship between HERV expression and tumor characteristics, we plotted HERV expression separately for each category of tumor microsatellite instability status and stage.
The tumor specific HERVs tended to be higher in microsatellite stable tumors with the notable exceptions of HERV-H Xp22.32 and HERV-H 20p11.23 (Supplementary Figure 4a).
No trend in expression level relative to tumor stage was apparent (Supplementary Figure 4b).
Finally, to determine the relationship between HERV expression and cytotoxic activity within tumors, we measured the correlation between HERV expression and the expression of six markers of cytotoxic activity.
The correlation was minimal for most HERVs, but HERV1-I 3p22.3 had a weakly negative association (Supplementary Figure 5).

To verify the consistency of associations between proviral expression and tissue phenotype and to validate the utility of tumor specific HERVs as potential biomarkers, we applied the same methods used in the discovery cohort to an independent cohort (independent cohort \#1) of 276 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 208 tumor samples (Supplementary Table 1).
Of the 150 HERVs moderately expressed (i.e. expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in independent cohort \#1.
Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure 1c).
As tumor classifiers in this cohort, most of the 11 tumor specific HERVs had weaker performance (Supplementary Figure 6).
However, HERV-H Xp22.32 and HERV1-I 3p22.3 maintained excellent discriminatory ability.
Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated as consistently tumor specific in the independent cohort (Table \ref{tab:val} and Supplementary Table 3).
Note that nominally significant (i.e. \emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered significant for the purpose of validation due to the small number of healthy samples in the independent cohort.

To confirm the accuracy of our bioinformatic and statistical methods, additional validation testing was performed on a third set of samples.
Testing in the discovery cohort relied on latent factor analysis to adjust for confounding factors across tissue phenotypes.
An alternative approach is to compare pairs of samples matched by subject.
This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for demographic, technical, and other possible confounders more effectively.
Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (independent cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples were highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1d).
The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.

\subsection*{Proviruses are expressed in healthy colorectal tissue}
Several HERV proviruses were expressed in tumor specific patterns, in accord with our hypothesis.
However, a surprising finding of our survey of unannotated retroviral transcripts in colorectal transcriptomes was the unexpectedly high level of expression of HERV elements in healthy samples.
The number of differentially expressed genes (\emph{P} adjusted < 0.05) with higher expression in healthy and tumor adjacent samples relative to tumor samples was 94 and 13, respectively.
The number of differentially expressed genes with lower expression in healthy and tumor adjacent samples relative to tumor samples was 28 and 14, respectively.
This result is demonstrated in Figure \ref{fig:dge_res}a,b (blue points).
A high number of differentially expressed genes with over-expression in healthy tissue was also found in independent cohort \#1 (data not shown).
